Predicting Tags for StackOverflow Posts

### 2013-02-22 stackExchange/writeups/iccm-2013/writeup.tex
foo
Former-commit-id: 237823efabeb3e20e84e3b99f6bb51000e21dc33 [formerly d0d7392f5f7c1e2459fdb352bf6373a8f274a7a5]
Former-commit-id: c80f84d1371d97d9a6ceb951bf2d169307611b3f

\usepackage{cogsci}
\usepackage{pslatex}
\usepackage{apacite}

@@

The Apr' 11 StackOverflow dataset \cite{DataDump2011} was used for this analysis.
The dataset contains \num{4945168} posts from \num{559803} unique users and \num{26176} unique tags.
The dataset is the entire collection of posts dated between Jul' 2008 (StackOverflow's creation) and Apr' 11.

The Apr' 11 StackOverflow dataset \cite{DataDump2011} was downloaded and imported into the MySQL relational database system.

A query was then used to build the post word, tag word co-occurrence matrix.
The result is a set of context word, tag, count for each of the \num{39223968} unique context word, tag combinations.

For a given title and body of a post, the model returns the activations for each possible tag.
The tag with the highest activation is the most likely tag to be associated with that post.
Tag activation is a function of four components:
The tag's base level activation (prior odds of being used),
contextual activation for words in the title (word co-occurrence weighting),
contextual activation for words in the body,
and an offset to equalize activation values for tags across all posts.
The equations that operationalize this process will be included in the Model section.

The base R glm function (with 'binomial' argument) was used to perform the logistic regression.
The parameters were optimized to best predict categorization (chosen tag, not chosen tag) from the four predictor variables:
The tag's prior occurrence frequency, word co-occurrence activation for title and body words, and activation offset for the post.
Further discussion and operationalization of the predictor variables is included in the Model section.

@@

\renewcommand{\arraystretch}{1.5}% Wider
\renewcommand{\tabcolsep}{.5mm}
\begin{table}[!ht]
  \begin{center} 
    \caption{The StackOverflow tag prediction model} 
    \label{sample-table} 
    \vskip 0.12in
    \begin{tabular}{ll} 
      \hline
      Common Name &  Equation \\
      \hline
      Activation & 		$A_{i} = B_{i} + \sum_{j\in T}^{ } W_{j} S_{ji} + \sum_{j\in B}^{ } W_{j} S_{ji} - O$ \\
      Base Level & 		$B_{i} = log \frac{p_{i}}{1-p_{i}}$ \\
      Strength Assoc. &		$S_{ji} = log \frac{p(i|j)}{p(i))} = log \frac{NN(j,i)}{N_{Row}(j)N_{Col}(i)}$ \\
      Attention Weight	& 	$W_{j} = WE_{j}$ \\
      Scaled Entropy & 		$E_{j} = 1-\frac{H_{j}}{H_{max}}$ \\
      Entropy & 		$H_{j} = -\sum_{i=1}^{N}p(i|j)log\left (  p(i|j) \right )$ \\
      Max Entropy & 		$H_{max} = maxH$ \\
      Offset & 			$O = \frac{1}{5}\sum_{i\in top 5}^{ } A_{i}$ \\
      \hline
    \end{tabular} 
  \end{center} 
\end{table}

@@

Base-level activation ($B_{i}$) is the log prior odds of observing tag $i$ in the world.
So the tag with the highest frequency (\emph{C\#}) would have the highest base-level activation.
Tag frequency drops off sharply for this dataset.
The top 1\% of tags based on frequency of use account for approximately 60\% of total tag instances.
This is why accounting for tag frequency (by including ACT-R's base-level activation component) is so important for the StackOverflow dataset.
Otherwise without context, the likelihood of choosing \emph{C\#} as a tag would be the same as choosing \emph{ora-00997} (one of the many tags with only a single occurrence of use).
One might argue that with context, it's unlikely that the words associated with \emph{ora-00997} will occur again, so pure context would be able to still often predict \emph{C\#}.
However, the ACT-R theory asserts that pure context is often not enough to overcome high base-level activations (e.g., it can be difficult to break a bad habit).

@@

Contextual activation can be thought of as the amount of odds adjustment to base levels, given the current context $j$.
Contextual activation has two components: strength of association ($S_{ji}$) and attentional weight ($W_{j}$).
Strength of association measures the relatedness of a contextual element $j$ (e.g., the word PHP in the title of a post) with a tag $i$ (e.g., the tag PHP5).

@@

Attentional weighting ($W_{j}$), can be thought of as the amount of attentional resources that are dedicated to contextual element $j$.
Total attentional weighting ($W$) is bound, most often set to 1, and reflects the fact that our attentional resources are limited.
For the StackOverflow model, the attentional weighting was left unbound and allowed to be calibrated by the logistic regression model.
This was done for two reasons:
First, the researcher was curious to see what value the logistic regression model would give for $W$.
Second, a non-ACT-R standard measure of attentional weighting was used ($E_{j}$), so it seemed appropriate to allow $W$ to vary alongside exploring this new weighting measure.

@@

During initial model testing, the attentional weighting for all words was equal.
That is initially, for a given post with $n$ title words, the attentional weighting $W_{j\in C}$ for all of those words was $W/n$.
The problem with this approach is that familiar stop words like 'the' and '?', that are often observed across all tags have the same attentional weighting as highly predictive words like 'PHP'.
Granted the $S_{ji}$'s across all tags for context words like 'the' are much lower (and more evenly distributed) than the few $S_{ji}$'s that spike for 'PHP' (e.g., the tag PHP).
So the amount of adjusted log odds from stop words is low even with equal attentional weighting (since their $S_{ji}$'s are low).
However, this still seemed problematic for two reasons:
First, because stop words like 'the' are observed with all tags, they are not highly predictive of any tag.
So one could argue that their presence is simply adding error in the model's tag activation calculation.
Second, because $W$ is held constant across posts, any attentional resources that are allocated to stop words like 'the' and '?' are taken away from being used for highly predictive words like 'PHP'.

It's also the case that there are many connections and commonalities between ACT-R's declarative memory retrieval equations and other non-ACT-R information retrieval theories,
and many of those theories use a combination of weighting schemes for contextual relevance and global importance of a particular chunk in context.
For example, \citeA{Dumais1991} shows that a measure of how likely term $j$ is in document $i$ is often the product of a local weighting $L_{ji}$ and global weighting $G_{j}$ measure.
This idea maps rather nicely to ACT-R's contextual weighting.
That is, terms are context words $j$, documents are tags $i$, local weighting is strength of association $S_{ji}$, and global weighting is attentional weighting $W_{j}$.

@@

For example, a stop word like 'the' will have an even distribution of co-occurrences across all tags.
This means that the marginal probabilities of 'the' with all tags will be roughly equal.
It can be shown that maximum entropy $H_{j}$ is reached when all marginal probabilities are equal.
This makes intuitive sense because if all marginal probabilities are equal, there is no information gained in observing the context word.
The scaled entropy measure $E_{j}$ transforms the entropy measure to a scale that ranges from 0, 1, where maximum entropy means lowest scaled entropy.
This means that a stop word like 'the' will have a scaled entropy measure close to 0, while a highly-predictive word like \emph{macro} will have a scaled entropy measure close to 1.
Going back to ACT-R's attentional weighting $W_{j}$, if $E_{j}$ is low then $W_{j}$ is low, meaning that the scaled entropy measure helps allocate the limited attentional resources to predictive contextual elements.

@@

An offset measurement (mean of the top five tags for the post) was used to equalize the top activated tags for each post.
This was done primarily to improve model fit.
To see why this is the case, suppose the following scenario:
The offset measurement is not used, and the model is asked to provide ten tags across ten posts.
The top ten tags for one of these posts have activations that are higher than activations for all other tags in the other nine posts.
The model would then select the ten tags from the single post as the most likely tags.
But this doesn't make intuitive sense, because there is a limit to the number of tags that people tend to use for each post (average number of tags per post is approximately 3).
So the offset measure aims to place all of these ten posts on the same level, so that asking the model to choose ten tags from ten posts will select roughly a single tag from each post.

@@

After training the model's spreading activation matrix and calibrating model parameters, the model can predict 67\% of tags correctly when asked to generate an average of one tag per post.

@@

Note how the base level activation and contextual activation combine to predict tags that are both frequently used and contextually relevant.
For example, note that the 'javascript' tag leverages a combination of contextual activation and prior activation to achieve a high total activation.
Alternatively, the 'svg' tag relies almost entirely on context to achieve high activation.
And note that the model is by no means perfect.
For example, 'html' is not tagged by the author, but the model ranks it as the second highest tag.
And the author's tag 'firefox' is only ranked 8th by the model.

@@

\begin{figure}[ht]
  \centering
  \resizebox{\linewidth}{!}{\includegraphics{ROC-cropped.pdf}}
  \caption{
    Model performance across the calibration and test dataset, as well as when specific model components are removed.
    The vertical dotted line represents the point where the model is asked to tag on average one tag per post.
  }
  \label{fig:ROC}
\end{figure}

@@

The Python NLTK documentation \cite{Bird2009} recommended to remove stop words from the analysis.
Stop words are low-predictor words, and the documentation listed out a short list that were commonly removed (e.g., 'the', 'and', 'or')
However it seemed somewhat arbitrary to explicitly list out a few stop words and then simply choose those as the words to remove.
Looking at the scaled entropy measure, it seems quite plausible that this measure of word predictiveness can also be used to identify stop words.
That is, words with scaled entropy close to zero are the stop words.
Looking at the lowest ten scaled-entropy words for this dataset, (does, anyone, 've, found, are, seems, has, been, for, support), using scaled entropy to determine stop words seems to work fairly well.

>>>>

\usepackage{cogsci}
\usepackage{pslatex}
\usepackage{apacite}
\usepackage[none]{hyphenat}

@@

The Apr' 11 StackOverflow dataset \cite{DataDump2011} was used for this analysis.
The dataset contains \num{4945168} posts from \num{559803} unique users and \num{26176} unique tags.
The dataset is the entire collection of posts dated between Jul' 2008 (StackOverflow's creation) and Apr' 11.
Post authors used approximately 3 tags on average per post.

@@

\renewcommand{\arraystretch}{1.5}% Wider
\renewcommand{\tabcolsep}{.5mm}
\begin{table}[!ht]
  \begin{center} 
    \caption{The StackOverflow tag prediction model} 
    \label{sample-table} 
    \vskip 0.12in
    \begin{tabular}{ll} 
      \hline
      Common Name &  Equation \\
      \hline
      Activation & 		$A_{i} = B_{i} + \sum_{j\in T}^{ } W_{j} S_{ji} + \sum_{j\in B}^{ } W_{j} S_{ji} - O$ \\
      Base Level & 		$B_{i} = log \frac{p_{i}}{1-p_{i}}$ \\
      Strength Assoc. &		$S_{ji} = log \frac{p(i|j)}{p(i))} = log \frac{NN(j,i)}{N_{Row}(j)N_{Col}(i)}$ \\
      Attention Weight	& 	$W_{j} = WE_{j}$ \\
      Scaled Entropy & 		$E_{j} = 1-\frac{H_{j}}{H_{max}}$ \\
      Entropy & 		$H_{j} = -\sum_{i=1}^{N}p(i|j)log\left (  p(i|j) \right )$ \\
      Offset & 			$O = \frac{1}{5}\sum_{i\in top 5}^{ } A_{i}$ \\
      \hline
    \end{tabular} 
  \end{center} 
\end{table}

@@

Base-level activation ($B_{i}$) is the log prior odds of observing tag $i$ in the world.
The log prior odds are based on both frequency and recency of use.
However we are are assuming that frequency of tag use does not change greatly between Jul' 08 and Apr' 11 (time period of dataset).
This allows us to compute log odds based on frequency only, as is indicated in Table \ref{sample-table}.
So the tag with the highest frequency (\emph{C\#}) would have the highest base-level activation.
Tag frequency drops off sharply for this dataset.
The top 1\% of tags based on frequency of use account for approximately 60\% of total tag instances.
This is why accounting for tag frequency (by including ACT-R's base-level activation component) is so important for the StackOverflow dataset.
Otherwise without context, the likelihood of choosing \emph{C\#} as a tag would be the same as choosing \emph{ora-00997} (one of the many tags with only a single occurrence of use).
One might argue that with context, it's unlikely that the words associated with \emph{ora-00997} will occur again, so pure context would be able to still often predict \emph{C\#}.
However, the ACT-R theory asserts that pure context is often not enough to overcome high base-level activations (e.g., it can be difficult to break a bad habit).

@@

Contextual activation can be thought of as the amount of odds adjustment to base levels, given the current context $j$.
Contextual activation has two components: strength of association ($S_{ji}$) and attentional weight ($W_{j}$).
Strength of association measures the relatedness of a contextual element $j$ (e.g., the word PHP in the title of a post) with a tag $i$ (e.g., the tag \emph{PHP5}).

@@

Attentional weighting ($W_{j}$), can be thought of as the amount of attentional resources that are dedicated to contextual element $j$.
Total attentional weighting ($W$) is bound, most often set to 1, and reflects the fact that attentional resources are limited.
For the StackOverflow model, the attentional weighting was left unbound and allowed to be calibrated by the logistic regression model.
This was done for two reasons:
First, the we were curious to see what value the logistic regression model would give for $W$.
Second, a non-ACT-R standard measure of attentional weighting was used ($E_{j}$), so it seemed appropriate to allow $W$ to vary alongside exploring this new weighting measure.

@@

During initial model testing, the attentional weighting for all words was equal.
That is initially, for a given post with $n$ title words, the attentional weighting $W_{j\in C}$ for all of those words was $W/n$.
The problem with this approach is that familiar stop words like ``the'' and ``?'', that are often observed across all tags have the same attentional weighting as highly predictive words like ``PHP''.
Granted the $S_{ji}$'s across all tags for context words like ``the'' are much lower (and more evenly distributed) than the few $S_{ji}$'s that spike for ``PHP'' (e.g., the tag \emph{PHP}).
So the amount of adjusted log odds from stop words is low even with equal attentional weighting (since their $S_{ji}$'s are low).
However, this still seemed problematic for two reasons:
First, because stop words like ``the'' are observed with all tags, they are not highly predictive of any tag.
So one could argue that their presence is simply adding error in the model's tag activation calculation.
Second, because $W$ is held constant across posts, any attentional resources that are allocated to stop words like ``the'' and ``?'' are taken away from being used for highly predictive words like ``PHP''.

It's also the case that there are many commonalities between ACT-R's declarative memory retrieval equations and other non-ACT-R information retrieval theories,
and many of those theories use a combination of weighting schemes for contextual relevance and global importance of a particular chunk in context.
For example, \citeA{Dumais1991} shows that a measure of how likely term $j$ is in document $i$ is often the product of a local weighting $L_{ji}$ and global weighting $G_{j}$ measure.
This idea maps rather nicely to ACT-R's contextual weighting.
That is, terms are context words $j$, documents are tags $i$, local weighting is strength of association $S_{ji}$, and global weighting is attentional weighting $W_{j}$.

@@

For example, a stop word like ``the'' will have an even distribution of co-occurrences across all tags.
This means that the marginal probabilities of ``the'' with all tags will be roughly equal.
It can be shown that maximum entropy $H_{j}$ is reached when all marginal probabilities are equal.
This makes intuitive sense because if all marginal probabilities are equal, there is no information gained in observing the context word.
The scaled entropy measure $E_{j}$ transforms the entropy measure to a scale that ranges from 0 to 1, where maximum entropy means lowest scaled entropy.
This means that a stop word like ``the'' will have a scaled entropy measure close to 0, while a highly-predictive word like \emph{macro} will have a scaled entropy measure close to 1.
Going back to ACT-R's attentional weighting $W_{j}$, if $E_{j}$ is low then $W_{j}$ is low, meaning that the scaled entropy measure helps allocate the limited attentional resources to predictive contextual elements.

@@

An offset measurement (mean of the top five tags for the post) was used to equalize the top activated tags for each post.
This was done primarily to improve model fit.
To see why this is the case, suppose the following scenario:
The offset measurement is not used, and the model is asked to provide ten tags across ten posts.
The top ten tags for one of these posts have activations that are higher than activations for all other tags in the other nine posts.
The model would then select the ten tags from the single post as the most likely tags.
But this doesn't make intuitive sense, because there is a limit to the number of tags that people tend to use for each post.
So the offset measure aims to place all of these ten posts on the same level, so that asking the model to choose ten tags from ten posts will select roughly a single tag from each post.

FIXME. The Apr' 11 StackOverflow dataset \cite{DataDump2011} was downloaded and imported into the MySQL relational database system.

For a given title and body of a post, the model returns the activations for each possible tag.
The tag with the highest activation is the most likely tag to be associated with that post.
Tag activation is a function of four components:
The tag's base level activation (prior odds of being used),
contextual activation for words in the title (word co-occurrence weighting),
contextual activation for words in the body,
and an offset to equalize activation values for tags across all posts.

A query was then used to build the post word, tag word co-occurrence matrix.
There are \num{39223968} unique post word, tag word combinations.
The result is a table of co-occurence counts for each of the combinations.

The base R glm function (with ``binomial'' argument) was used to perform the logistic regression.
The parameters were optimized to best predict categorization (chosen tag, not chosen tag) from the four predictor variables:
The tag's prior occurrence frequency, word co-occurrence activation for title and body words, and activation offset for the post.
Further discussion and operationalization of the predictor variables is included in the Model section.

@@

@@

Note how the base level activation and contextual activation combine to predict tags that are both frequently used and contextually relevant.
For example, note that the \emph{javascript} tag leverages a combination of contextual activation and prior activation to achieve a high total activation.
Alternatively, the \emph{svg} tag relies almost entirely on context to achieve high activation.
And note that the model is by no means perfect.
For example, \emph{html} is not tagged by the author, but the model ranks it as the second highest tag.
And the author's tag \emph{firefox} is only ranked 8th by the model.

@@

\begin{figure}[ht]
  \centering
  \resizebox{\linewidth}{!}{\includegraphics{ROC-cropped.pdf}}
  \caption{
    Model performance across the calibration and test dataset, as well as when specific model components are removed.
    The y-axis represents model performance (one is perfect), and the x-axis is the proportion of tags that the model generates compared to the number of tags used by the authors of the posts (one is equal number).
    The vertical dotted line represents the point where the model is asked to tag on average one tag per post.
  }
  \label{fig:ROC}
\end{figure}

@@

The Python NLTK documentation \cite{Bird2009} recommended to remove stop words from the analysis.
Stop words are low-predictor words, and the documentation listed out a short list that were commonly removed (e.g., ``the'', ``and'', ``or'')
However it seemed somewhat arbitrary to explicitly list out a few stop words and then simply choose those as the words to remove.
Looking at the scaled entropy measure, it seems quite plausible that this measure of word predictiveness can also be used to identify stop words.
That is, words with scaled entropy close to zero are the stop words.
Looking at the lowest ten scaled-entropy words for this dataset, (``does'', ``anyone'', ``'ve'', ``found'', ``are'', ``seems'', ``has'', ``been'', ``for'', ``support''),
using scaled entropy to determine stop words seems to work fairly well.
